---
editor_options: 
  chunk_output_type: console
---

# Abundance vs. acoustic detections  

In this script, we correlate and regress the abundance from point count data with the number of acoustic detections from the paired audio dataset, as classified by BirdNET.

Here, abundance corresponds to the total number of individuals of a species detected for point count data. 

## Install necessary libraries
```{r}
library(tidyverse)
library(dplyr)
library(stringr)
library(vegan)
library(ggplot2)
library(scico)
library(data.table)
library(extrafont)
library(ggstatsplot)
library(viridis)   
library(ggrepel)  
```

## Load existing data
```{r}
datSubset <- read.csv("results/datSubset.csv")
```

## Abundance and acoustic detection comparisons
```{r}
# estimate total abundance of all species for each site by season
abundance <- datSubset %>%
  filter(dataType == "point_count") %>%
  group_by(year_season, scientific_name,
           common_name, gridID) %>%
  summarise(abundance_pc = sum(number)) %>%
  ungroup()

# estimate total acoustic detections per site and season for each species
detections <- datSubset %>%
  filter(dataType == "acoustic_data") %>%
  group_by(year_season, scientific_name,
           common_name, gridID)  %>% 
  summarise(detections_aru = sum(number)) %>%
  ungroup()

# create a single dataframe
data <- full_join(abundance, detections)%>%
  replace_na(list(abundance_pc = 0, detections_aru = 0))

# we subset species to only include data with a minimum number of abundance and acoustic detections 
spp_subset <-  data %>%
  group_by(common_name, year_season) %>%
  summarise(abundance_pc = sum(abundance_pc), detections_aru = sum(detections_aru)) %>%
  ungroup() %>%
  filter(abundance_pc >=30 & detections_aru >= 30)

summer_subset <-  spp_subset %>%
  filter(year_season == "2024 Summer")
winter_subset <-  spp_subset %>%
  filter(year_season == "2024 Winter")

# subset data by season
# we have 11 species in total for summer and 12 for winter
summer <- data %>%
  filter(year_season == "2024 Summer") %>%
  filter(common_name %in% summer_subset$common_name)
winter <- data %>%
  filter(year_season == "2024 Winter") %>%
  filter(common_name %in% winter_subset$common_name)

# visualization by season across species for summer
fig_summer_abundance_detections_cor <- 
  ggscatterstats(
  data = summer,
  y = abundance_pc,
  x = detections_aru,
  type = "r",
  ylab = "Abundance (point-count)",
  xlab = "Acoustic detections (ARU)",
  ggplot.component = list(theme(text = element_text(family = "Century Gothic", size = 15, face = "bold"),plot.title = element_text(family = "Century Gothic",
      size = 18, face = "bold"),
      plot.subtitle = element_text(family = "Century Gothic", 
      size = 15, face = "bold",color="#1b2838"),
      axis.title = element_text(family = "Century Gothic",
      size = 15, face = "bold"))))

ggsave(fig_summer_abundance_detections_cor, filename = "figs/fig_abundance_vs_acousticDetections_correlations_summer.png", width = 14, height = 10, device = png(), units = "in", dpi = 300)
dev.off() 

# visualization by season across species for winter
fig_winter_abundance_detections_cor <- 
  ggscatterstats(
  data = winter,
  y = abundance_pc,
  x = detections_aru,
  type = "r",
  ylab = "Abundance (point-count)",
  xlab = "Acoustic detections (ARU)",
  ggplot.component = list(theme(text = element_text(family = "Century Gothic", size = 15, face = "bold"),plot.title = element_text(family = "Century Gothic",
      size = 18, face = "bold"),
      plot.subtitle = element_text(family = "Century Gothic", 
      size = 15, face = "bold",color="#1b2838"),
      axis.title = element_text(family = "Century Gothic",
      size = 15, face = "bold"))))

ggsave(fig_winter_abundance_detections_cor, filename = "figs/fig_abundance_vs_acousticDetections_correlations_winter.png", width = 14, height = 10, device = png(), units = "in", dpi = 300)
dev.off() 
```

## Regressions between abundance and acoustic detections  

While it is relevant to examine correlations between abundance (as estimated from point count data) and acoustic detections (as estimated from acoustic data), we will run robust regressions to ask one predicts the other.  
```{r}
## I am borrowing code from Mikula et al. 2020 below
## The data and scripts from their manuscript can be found here:
## https://osf.io/fa9ky/

# splitting by seasons
summer <- setDT(summer)
winter <- setDT(winter)

# extract t-value
summer[,  t_value := summary(lm(abundance_pc~detections_aru))$coefficients[6], by = common_name] 
winter[,  t_value := summary(lm(abundance_pc~detections_aru))$coefficients[6], by = common_name] 

# extract slope
summer[,  slope := lm(abundance_pc~detections_aru)%>% coef()%>% nth(2), by = common_name] 
winter[,  slope := lm(abundance_pc~detections_aru)%>% coef()%>% nth(2), by = common_name] 

# extract pearson's correlation
summer[,  pearson := cor(abundance_pc, detections_aru), by = common_name] 
winter[,  pearson := cor(abundance_pc, detections_aru), by = common_name] 

# extract adjusted r squared
summer[,  r_sq := summary(lm(abundance_pc~detections_aru))$adj.r.squared, by = common_name] 
winter[,  r_sq := summary(lm(abundance_pc~detections_aru))$adj.r.squared, by = common_name] 

# create a column with the direction of the slope (whether it is positive or negative), which can be referred to later while plotting    
summer[, slope_dir := ifelse(slope >0, '+', '-')]
winter[, slope_dir := ifelse(slope >0, '+', '-')]

paste("Positive regressions:",length(unique(summer$common_name[summer$slope_dir %in% c('+')])))
# 16 species had a positive regression/slope value in summer
paste("Positive regressions:",length(unique(winter$common_name[winter$slope_dir %in% c('+')])))
# 21 species had a positive regression/slope value in summer

## visualization for summer
fig_summer_abund_detec_reg <- ggplot(summer, aes(x = detections_aru,
                                              y = abundance_pc)) +
  geom_point(color = "#9CC3D5",size = 1.2) +
  geom_smooth(data = summer, aes(group = common_name,
                                     color = slope_dir), 
              method = 'lm', se = FALSE, 
              linewidth = 0.7) +
  scale_color_manual(values=c("#1B9E77", "#D95F02")) +
  labs(x="\nAcoustic detections (from acoustic data)", 
       y="Abundance (from point count data)\n") + 
  theme_bw() +
  annotate("text", y=10, x=40, 
           label= "Slope:", col = "grey30", size = 12) + 
  annotate("text", y=8, x=40, 
           label= "+", col = "#D95F02", size = 12) + 
  annotate("text", y = 7, x=40, 
           label = "-", col = "#1B9E77", size = 12)+
  theme(text = element_text(family = "Century Gothic", size = 18, face = "bold"),plot.title = element_text(family = "Century Gothic",
      size = 18, face = "bold"),
      plot.subtitle = element_text(family = "Century Gothic", 
      size = 15, face = "bold",color="#1b2838"),
      axis.title = element_text(family = "Century Gothic",
      size = 18, face = "bold"),
      legend.position = "none")

ggsave(fig_summer_abund_detec_reg, filename = "figs/fig_abundance_vs_acousticDetections_regressions_summer.png", width = 14, height = 16, device = png(), units = "in", dpi = 300)
dev.off() 

## visualization for winter
fig_winter_abund_detec_reg <- ggplot(winter, aes(x = detections_aru,
                                              y = abundance_pc)) +
  geom_point(color = "#9CC3D5",size = 1.2) +
  geom_smooth(data = winter, aes(group = common_name,
                                     color = slope_dir), 
              method = 'lm', se = FALSE, 
              linewidth = 0.7) +
  scale_color_manual(values=c("#1B9E77", "#D95F02")) +
  labs(x="\nAcoustic detections (from acoustic data)", 
       y="Abundance (from point count data)\n") + 
  theme_bw() +
  annotate("text", y=10, x=40, 
           label= "Slope:", col = "grey30", size = 12) + 
  annotate("text", y=8, x=40, 
           label= "+", col = "#D95F02", size = 12) + 
  annotate("text", y = 7, x=40, 
           label = "-", col = "#1B9E77", size = 12)+
  theme(text = element_text(family = "Century Gothic", size = 18, face = "bold"),plot.title = element_text(family = "Century Gothic",
      size = 18, face = "bold"),
      plot.subtitle = element_text(family = "Century Gothic", 
      size = 15, face = "bold",color="#1b2838"),
      axis.title = element_text(family = "Century Gothic",
      size = 18, face = "bold"),
      legend.position = "none")

ggsave(fig_winter_abund_detec_reg, filename = "figs/fig_abundance_vs_acousticDetections_regressions_winter.png", width = 14, height = 16, device = png(), units = "in", dpi = 300)
dev.off() 

# extract the slope, t_value, pearson correlation and the adjusted r square
lm_output_summer <- summer %>%
  dplyr::select(common_name, t_value, slope, pearson, slope_dir,r_sq) %>% distinct() %>%
  mutate(year_season = "2024 Summer")
lm_output_winter <- winter %>%
  dplyr::select(common_name, t_value, slope, pearson, slope_dir,r_sq) %>% distinct() %>%
  mutate(year_season = "2024 Winter")

# write the values to file
lm_output <- bind_rows(lm_output_summer, lm_output_winter)
write.csv(lm_output, "results/abundance-acousticDetections-regressions.csv",
          row.names = F)
```

## Plotting species-specific regression plots  

```{r}
# visualization for summer
plots <- list()

for(i in 1:length(unique(summer$common_name))){
  
  # extract species scientific name
  a <- unique(summer$common_name)[i]
  
  # subset summer for plotting
  for_plot <- summer[summer$common_name==a,]
  
  # create plots
  plots[[i]] <- ggplot(for_plot, aes(x = detections_aru,
                                    y = abundance_pc)) +
  geom_point(color = "#00496FFF",size = 1.2) +
  geom_smooth(aes(color = "#D95F02"),
    method = 'lm', se = TRUE, 
              linewidth = 0.7) +
    labs(title = paste0(a,"  ","r_sq = ", signif(for_plot$r_sq, digits = 2), "  ", paste0("slope = ",signif(for_plot$slope, digits = 4))),
      x="\nAcoustic Detections (from acoustic data)", 
       y="Abundance (from point count data)\n") +
    theme_bw() +
  theme(text = element_text(family = "Century Gothic", size = 18, face = "bold"),plot.title = element_text(family = "Century Gothic",
      size = 18, face = "bold"),
      plot.subtitle = element_text(family = "Century Gothic", 
      size = 15, face = "bold",color="#1b2838"),
      axis.title = element_text(family = "Century Gothic",
      size = 18, face = "bold"),
      legend.position = "none")
}

# plot and save as a single pdf
cairo_pdf(
  filename = "figs/abundance-acousticDetections-by-species-regressions-summer.pdf",width = 13, height = 12,
  onefile = TRUE
)
plots
dev.off() 

# visualization for winter
plots <- list()

for(i in 1:length(unique(winter$common_name))){
  
  # extract species scientific name
  a <- unique(winter$common_name)[i]
  
  # subset winter for plotting
  for_plot <- winter[winter$common_name==a,]
  
  # create plots
  plots[[i]] <- ggplot(for_plot, aes(x = detections_aru,
                                    y = abundance_pc)) +
  geom_point(color = "#00496FFF",size = 1.2) +
  geom_smooth(aes(color = "#D95F02"),
    method = 'lm', se = TRUE, 
              linewidth = 0.7) +
    labs(title = paste0(a,"  ","r_sq = ", signif(for_plot$r_sq, digits = 2), "  ", paste0("slope = ",signif(for_plot$slope, digits = 4))),
      x="\nAcoustic Detections (from acoustic data)", 
       y="Abundance (from point count data)\n") +
    theme_bw() +
  theme(text = element_text(family = "Century Gothic", size = 18, face = "bold"),plot.title = element_text(family = "Century Gothic",
      size = 18, face = "bold"),
      plot.subtitle = element_text(family = "Century Gothic", 
      size = 15, face = "bold",color="#1b2838"),
      axis.title = element_text(family = "Century Gothic",
      size = 18, face = "bold"),
      legend.position = "none")
}

# plot and save as a single pdf
cairo_pdf(
  filename = "figs/abundance-acousticDetections-by-species-regressions-winter.pdf",width = 13, height = 12,
  onefile = TRUE
)
plots
dev.off() 
```

